\chapter{Debugging}
\label{Debugging}

As you begin to develop your kernel, you will undoubtedly find problems in the code you have written. We have collected some techniques here which we found useful or which are largely undocumented elsewhere because they are fairly advanced or are specific to working on Weenix in particular.

\section{Beginning Debugging}

These are the simplest debugging techniques, which can be used at any stage of the project. They require the least amount of learning, but they also are not nearly as powerful as the debugging techniques listed later on.

\subsection{Printing}

By far the easiest technique for debugging is printing things out during the execution of your program. Of course, printing inside the kernel requires significant overhead (as you will learn when you write the TTY driver). Luckily, inside the support code we have provided printing methods which use the serial port of the virtual machine to print to a terminal outside of the computer. You can think of this as being like a printer connected to the serial port if you wish.  The way to use this is through the \texttt{dbg} or \texttt{dbgq} macro.
\begin{verbatim}
#define dbg(dbgmodes, printfargs)
#define dbgq(dbgmodes, printfargs)\end{verbatim}
The difference between the two is that \texttt{dbg} additionally displays the file and line number information describing where the debug statement is located, but \texttt{dbgq} displays only the debugging message.

Debug messages are organized into many debug modes, each of which can be separately hidden or color-coded in the debug output. You can find a list of debug modes in \texttt{kernel/include/util/debug.h}. The string names next to each mode are used to configure which debug modes are displayed (the default can be set in the Weenix configuration file and updated while Weenix is being debugged via the dbg command), and there is also a special string name which refers to all debug modes. For example:
\begin{verbatim}
dbg(DBG_THR, "Creating a kernel thread "
             "with stack at 0x%p\n", stack);\end{verbatim}
In the snippet above, \texttt{DBG\_THR} is the mode macro for the threading subsystem. This debug message would print something like:

\begin{verbatim}
kthread.c:123 kthread_stack_create(): Creating a kernel thread
with stack at 0x804800\end{verbatim}
It is also possible to define a debugging message which is part of multiple modes by taking their bitwise or.
\begin{verbatim}
dbg(DBG_THR | DBG_VM, "Creating a kernel thread with stack at "
                      "0x%p\n", stack);\end{verbatim}
It's not hard to add your own debugging modes, either. This is helpful if you decide to implement extra features and would like an easy way to separate their debug output from the normal subsystems without turning all of the debug modes off.

\subsection{Printing Using Info Functions}

There are several info functions in the kernel which are provided exclusively for visualizing information in the debug console. To call one of these, use \texttt{dbginfo()}, like so:
\begin{verbatim}dbginfo(DBG_VMMAP, vmmap_mapping_info, curproc->p_vmmap);\end{verbatim}
You can substitute in other functions besides \texttt{vmmap\_mapping\_info} - there are a bunch of functions ending with ``\texttt{\_info}'' in Weenix which can be passed here.

\subsection{Using Assertions}

Another simple but widely-applicable approach is to intentionally crash your program when some condition is not met, which is known as ``asserting'' that condition's truth. This allows you to know what the failure point was, which is sometimes helpful for figuring out what caused the failure in the first place, but mostly it's helpful just for letting you know that there was a failure instead of failing silently.

We provide assert functionality with the \texttt{KASSERT()} macro, which tests the condition you pass it and prints out an error message on crash that tells you the condition that failed and where in the codebase the \texttt{KASSERT()} was. A trick you may find helpful is that any string will be evaluated to true in C (because it's just a pointer) so you can do things like this:
\begin{verbatim}
KASSERT(a == 1 && "A should be equal to 1");\end{verbatim}
This may be more helpful than not using a string, since the message will be printed when the assertion fails. Of course, you can also just put a comment next to the assertion in the code.

Because assertions are helpful for checking things that should always be true, we recommend placing them at the beginning of any function whose arguments must be of a certain form to ensure that the caller is checking and filtering all error cases correctly (this is especially helpful for syscalls). Don't confuse this with error-checking that you need to do, though -- if you use assertions to check for error conditions you are supposed to handle, you will needlessly cause your kernel to panic!

\section{Intermediate Debugging}

Although the techniques above are very useful, it will quickly become apparent that a more robust way to debug the kernel is frequently necessary. Luckily, debuggers provide us with an excellent way to do this. \texttt{gdb} is the definitive debugger for C code on Unix systems, and the rest of this appendix will center on its usage. Most of the following information is applicable to all stages of Weenix development, except where noted.

\subsection{Prerequisites}
If you have never used \texttt{gdb} before, we recommend finding \hyperref{http://www.unknownroad.com/rtfm/gdbtut/gdbtoc.html}{}{}{RMS's (Richard Stallman's) \texttt{gdb} Debugger Tutorial} online and reading it. Before you read more advanced or Weenix-specific debugging tips, make sure you can use the following \texttt{gdb} features (most important ones listed first, more specialized ones later).
\begin{itemize}
    \item Compiling with symbols.
    \item Breakpoints, stepping, continuing.
    \item Viewing and traversing the call stack.
    \item Inspecting variables.
    \item Inspecting the contents at a particular memory address.
    \item Using watchpoints.\footnote{Watchpoints do not work with all the simulators which Weenix will run on.}
    \item Conditional breakpoints.\footnote{Conditional breakpoints do not work with all simulators which Weenix will run on. To get similar functionality, add an \texttt{if} statement to your code with the condition you care about and set a breakpoint inside there.}
    \item Inspecting the contents of registers.
\end{itemize}
Once you understand these basics, you should be able to debug pretty much any simple user-level program that you have the code for. However, there are a few more specific topics that are helpful in some cases.

\subsection{Debugging Multiple Threads}

Although this does not relate directly to Weenix, you will frequently be debugging multithreaded programs. Debugging in the presence of multiple threads is usually the same as debugging single-threaded programs, but the place where it differs is in cases like deadlock. If you are debugging a program in \texttt{gdb} and it deadlocks, you can hit \texttt{Ctrl-C} to pause the program. Then, you can inspect the stack trace as usual. If you need to check the other threads (perhaps the first one you were given wasn't the one in a deadlock, or maybe you want to see what other threads are contributing to the deadlock) you can see a list of threads using ``\texttt{info threads}'' and can switch to thread \texttt{N} by typing ``\texttt{thread N}'' (substitute the real thread number for \texttt{N} in that command).

Another note about multiple threads (particularly if you might be canceling them) is that you must be careful what calls you make during critical sections of code. For instance, if you want to ensure that a call to cancel some thread doesn't take effect until after a certain section of code is complete, there is a set of standard library calls which make system calls which might act as a cancellation point. Look in the \texttt{man} page for \texttt{pthread\_cancel()} for a list of these library calls. Note that \texttt{printf()} is one of them, so you might want to use \texttt{gdb} to debug instead of printing out messages to tell you when certain events happen.

\subsection{Using the Weenix \texttt{gdb} Scripts}

Because Weenix has been debugged using \texttt{gdb} for some time, there are a few Python scripts which run as custom commands inside the debugger which you can use to help debug your OS. These are all given under the command ``\texttt{kernel}'', and you can find more detailed information about the commands available by running ``\texttt{help kernel}'' inside of \texttt{gdb}, but here are a few of the highlights.
\begin{itemize}
    \item To turn kernel memory checking on, add the line ``\texttt{set kernel memcheck on}'' to the beginning of \texttt{init.gdb}. This allows you to run \texttt{kernel page} and \texttt{kernel slab} at shutdown, which tell you how much memory has not been cleaned up. Note that turning memory checking on may slow Weenix down a little bit, and that a handful of memory segments simply cannot be freed (page tables and stacks are common culprits).
    \item You can access the debug info functions mentioned above through the \texttt{kernel info} command.
    \item \texttt{kernel list} will tell you all the elements of a linked list if you used the macro list implementation.
    \item \texttt{kernel proc} will give you a list of all live processes.
\end{itemize}
Note that these require a certain version of the Python language and \texttt{gdb} to run; an error will be printed at the beginning of your \texttt{gdb} sessions if your versions are not compatible.

\subsection{Disassembling a Program}

If you need to know what exact instructions are being run or where exactly in your code you are executing, you should probably disassemble your program. The easiest way to do this inside \texttt{gdb} is to use the \texttt{disas} command, although you can also use a command like \texttt{x/100i \$eip} to print the next 100 instructions starting at the address pointed to by your instruction pointer. You can also use the \texttt{objdump} tool (separate from \texttt{gdb}) to disassemble the entire contents of an executable.

\section{Advanced Debugging Techniques}
Finally, there are some times when the approaches above are just not enough. These tips are for Weenix in particular, although they can easily be adapted for use in debugging many other kernels as well. These techniques will only be relevant when working on the virtual memory system of Weenix, since they are mainly for debugging problems in userland processes.

\subsection{Debugging a Page Fault}

The most obvious thing to do when debugging a page fault is to look at the address the fault is happening on. However, this doesn't tell you the context under which the page fault was generated, such as the stack or the section of code that was running. To find these, the easiest way is to inspect the stored context of the process that generated the pagefault. This will allow you to see the address of the instruction which was running (and perhaps more importantly, the name of its containing function), the stack pointer, and any other registers which might be needed to figure out what the processor was doing. This trick comes in especially handy when you also load in the symbol file for the user-level process, as the next section will show.

\subsection{Debugging processes from the kernel debugger}

It is a little bit like black magic, but you can actually debug user processes from inside the kernel debugger if you load in the symbol file for the user process into the correct location in memory. To find this information, you can use \texttt{objdump} (in your ordinary shell) to get some information about the user process.
\begin{verbatim}
$ objdump --headers --section=".text" user/simple/hello

user/simple/hello:  file format elf32-i386

Sections:
Idx Name          Size      VMA       LMA       File off  Algn
  6 .text         00000064  08048208  08048208  00000208  2**2
                  CONTENTS, ALLOC, LOAD, READONLY, CODE\end{verbatim}
The relevant information that this gives us is the starting address of the text section, \texttt{0x08048208} (the \texttt{VMA} column). So, we can open up Weenix with \texttt{gdb} attached, and add in the relevant symbol file for our userland process.
\begin{verbatim}
Breakpoint 3, bootstrap (arg1=0, arg2=0x0) at main/kmain.c:121
121     {
(gdb) add-symbol-file user/simple/hello 0x08048208
add symbol table from file "user/simple/hello" at
        .text_addr = 0x8048208
(y or n) y
Reading symbols from user/simple/hello...done.
(gdb) b main
Breakpoint 4 at 0x8048208: file ./hello.c, line 12.
(gdb) c
Continuing.
...
Breakpoint 4, main (argc=1, argv=0x8047eec) at ./hello.c:12
12      {
(gdb) list
7
8       #include <unistd.h>
9       #include <fcntl.h>
10
11      int main(int argc, char **argv)
12      {
13              open("/dev/tty0", O_RDONLY, 0);
14              open("/dev/tty0", O_WRONLY, 0);
15
16              write(1, "Hello, world!\n", 14);\end{verbatim}
Now, you should be able to set breakpoints in the userland process.

It is worth noting, however, that this does not prevent you from also putting breakpoints in kernel code. \texttt{gdb} is intentionally dumb about how breakpoints work - whenever your instruction pointer reaches the specified address, \texttt{gdb} will pause, no matter what symbol files you've added - so since the text of your kernel and the user process are both loaded, you can place breakpoints in either one.

\subsection{\texttt{gdb} and \texttt{DYNAMIC}}

With dynamic linking enabled, it becomes a step more difficult to debug userland processes from the kernel debugger, but certainly not impossible. The main issue is that there is a bunch more code in the \texttt{ld-weenix} shared library that you won't be able to debug unless you tell \texttt{gdb} to load the symbol file. The best way we know of to do this is to print out the memory map of the process you're trying to debug and make an educated guess about which region might correspond to \texttt{ld-weenix}. (It should be a shared region with execute permissions.) Then, load the debugging symbols starting at the address which corresponds to the beginning of that region.
